Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-30 transmission.tex
Dani comments

In this chapter, I use the estimate of the distribution of the duration of RT-PCR positivity from the combination of information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
I will show that the CIS data are sufficient, demonstrating the value of surveys of this type for pandemic surveillance.

I take two approaches, backcalculation and a mechanistic model, both previously discussed in \cref{E-inc-prev:sec:infection-process}.

I start with the backcalculation (\cref{backcalc}), based on \cref{E-inc-prev:eq:EPt}, which enables estimation of incidence within (sub)populations of England.
The approach makes few assumptions about how transmission occurs, decreasing the potential to introduce bias and can also easily incorporate arbitrary covariates to adjust for non-response bias.
The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.

@@

\section{Backcalculation} \label{backcalc}

@@

Specifying $F_Z$ should consider the need to provide structure to stabilise the problem (see \cref{E-intro:sec:disease-model}).
In the context of the CIS, it needs to allow borrowing information across time and/or strata (see \cref{E-biology-data:sec:MRP}).

The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{inc-prev:eq:forward-substitute}.
At a high-level, the statistical model is:
\begin{align}
    \vec{P}  &\dist F_P \\
    \rho &\dist F_\rho \\
    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
\end{align}
where $F_P$ is analogous to $F_Z$ in providing a joint prior for the prevalence in each stratum and day combination, and all other notation is as in the incidence-inferred approach.
The first step in the approach is to calculate the posterior distribution for $\vec{P}$.

@@

The incidence-inferred approach is more parsimonious in being a single step procedure.
However, computational challenges make makes mean I favour the prevalence-inferred approach.

It will be convenient to define a variable equating to the quantity initially estimated.
Let $\pi_{i,t} = Z_{i,t}$ under the inference-inferred approach and $\pi_{i,t} = P_{i,t}$ under the prevalence-inferred approach.

@@

The logit transformation also allows property 4 to be conveniently satisfied using an additive model for $g_{\vec{\varphi}}$.
An additive model on the logit scale is approximately multiplicative on the natural scale for small values of $\pi_{i,t}$ (which will be common).

This form of model includes the MRP prevalence model in \cref{E-biology-data:sec:MRP}.

@@

During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
There is weak evidence that when incidence starts increasing is correlated with the timing of the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for (A) all of England and (B) its component regions.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}
\begin{figure}
    \includegraphics{transmission/backcalc-alpha}
    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
        Linear regression line shown.
    }
    \label{transmission:fig:backcalc-alpha}
\end{figure}

The incidence proportion was highest among the 12--16 and 17--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
Except the 17--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information in the data to overcome the prior which favoured this situation.
The incidence proportion in the 17--25 age group decreases faster than other age groups during November.
There are many plausible explanations for this effect.
The higher incidence in this age group prior to November means it could be an immunity effect.
Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
Many other explanations are possible, and further work is required to determine causality.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
    \caption[Incidence estimated using backcalculation by age group]{%
        Incidence proportion for each age group, on both natural and log scales.
    }
    \label{transmission:fig:backcalc-ages}
\end{figure}

An edge effect, a rapid rise in incidence between days one and two, is visible (see \cref{transmission:fig:backcalc-start-effect}).
This is due to the incorrect assumption that incidence is constant in the past.
Changing this assumption (\eg to exponential growth) modifies the estimated value on the first day, but the rest of the series is insensitive to this change (not shown).
Therefore, the first day's estimate is unreliable, and has been excluded from all other results.
However, the rest of the series is robust to this assumption.
\begin{figure}
    \centering \includegraphics{transmission/backcalc-start-effect}
    \caption[Edge effects in backcalculation method]{%
        First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
    }
    \label{transmission:fig:backcalc-start-effect}
\end{figure}

@@

\Cref{transmission:fig:compare-regions} compares the regional incidence estimates between the two models in this chapter.
Generally, the estimates agree with the SEIR model having much narrower credible intervals.
The largest discrepancy is that the backcalculation model estimates a second peak in mid-January in several regions (most clearly in London, North West, and the East of England).
However, except in London, the credible intervals still overlap.
This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.
\begin{figure}
    \centering \includegraphics{transmission/compare-regions}
    \caption[Comparing each models estimate of regional incidence.]{%
        Incidence proportion in England and each region estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-regions}
\end{figure}

@@

The backcalculation method benefits from being able to be implemented in a fast, approximate inference package (R-INLA) that easily scales to many parameters, including random effects.
Therefore, it is far more flexible and can easily include time-varying effects, ethnicity, etc.
The SEIR model is limited by the inference procedure and the requirement to solve an ODE to evaluate the likelihood.
To incorporate a more flexible model, and reduce its bias, a more efficient inference procedure would be required.

Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev,SEIR:fig:prev-appendix}).
This is because the model is not flexible enough to capture the true age distribution of infections.
The overestimation of prevalence in the 12--16 age group and underestimation in the 17--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
\begin{figure}
    \centering \includegraphics{transmission/compare-NE}
    \caption[Comparing each models estimate of North East incidence proportion by age.]{%
        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-NE}
\end{figure}

>>>>

In this chapter, I use the estimate of the distribution of the duration of RT-PCR positivity based on combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
I will show that the CIS data are sufficient, demonstrating the value of surveys of this type for pandemic surveillance.

I take two approaches, one phenomenological and one mechanistic, both previously discussed in \cref{E-inc-prev:sec:infection-process}.

I start with the phenomenological approach (\cref{backcalc}), based on \cref{E-inc-prev:eq:EPt}, which enables estimation of incidence within (sub)populations of England.
The approach makes few assumptions about how transmission occurs, decreasing the potential to introduce bias and can also easily incorporate arbitrary covariates to adjust for non-response bias.
The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.

@@

\section{Phenomenological approach} \label{backcalc}

@@

Specifying $F_Z$ should consider the need to provide structure to stabilise the problem (see \cref{E-intro:sec:disease-model}).
In the context of the CIS, it needs to allow borrowing information across time and/or strata (see \cref{E-biology-data:sec:MRP}), which is required because for some stratum/time combinations the data become very sparse.

The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
First, it estimates prevalence using the statistical model from \cref{E-biology-data:sec:MRP}.
Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{inc-prev:eq:forward-substitute}.
At a high-level, the statistical model is:
\begin{align}
    \vec{P}  &\dist F_P \\
    \rho &\dist F_\rho \\
    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
\end{align}
where $F_P$ is analogous to $F_Z$ in providing a joint prior for the prevalence in each stratum and day combination, and all other notation is as in the incidence-inferred approach.
The first step in the approach is to calculate the posterior distribution for $\vec{P}$.

@@

The incidence-inferred approach is more parsimonious in being a single step procedure.
However, computational challenges make the prevalence-inferred approach more appealing.

It will be convenient to define a variable equating to the quantity initially estimated: let $\pi_{i,t} = Z_{i,t}$ under the inference-inferred approach and $\pi_{i,t} = P_{i,t}$ under the prevalence-inferred approach.

@@

The logit transformation also allows property 4 to be conveniently satisfied using an additive model for $g_{\vec{\varphi}}$.
An additive model on the logit scale is approximately multiplicative on the natural scale for small values of $\pi_{i,t}$, which will be common.

This form of the model includes the MRP prevalence model in \cref{E-biology-data:sec:MRP}.

@@

During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
There is weak evidence that when incidence starts increasing is correlated with the timing of the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using the phenomenological approach by region]{%
        Incidence proportion for (A) all of England and (B) its component regions.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}
\begin{figure}
    \includegraphics{transmission/backcalc-alpha}
    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
        Linear regression line shown.
    }
    \label{transmission:fig:backcalc-alpha}
\end{figure}

The incidence proportion was highest among the 12--16 and 17--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
Except the 17--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information in the data to overcome the prior which favoured this situation.
The incidence proportion in the 17--25 age group decreases faster than other age groups during November.
There are many plausible explanations for this effect.
The higher incidence in this age group prior to November means it could be an immunity effect.
Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
Many other explanations are possible, and further work is required to determine causality.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
    \caption[Incidence estimated using the phenomenological approach by age group]{%
        Incidence proportion for each age group, on both natural and log scales.
    }
    \label{transmission:fig:backcalc-ages}
\end{figure}

An edge effect, a rapid rise in incidence between days one and two, is visible (see \cref{transmission:fig:backcalc-start-effect}).
This is due to the incorrect assumption that incidence is constant in the past.
Changing this assumption (\eg to exponential growth) modifies the estimated value on the first day, but the rest of the series is insensitive to this change (not shown).
Therefore, the first day's estimate is unreliable, and has been excluded from all other results.
However, the rest of the series is robust to this assumption.
\begin{figure}
    \centering \includegraphics{transmission/backcalc-start-effect}
    \caption[Edge effects in the phenomenological approach method]{%
        First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
    }
    \label{transmission:fig:backcalc-start-effect}
\end{figure}

@@

\Cref{transmission:fig:compare-regions} compares the regional incidence estimates between the two models in this chapter.
Generally, the estimates agree with the SEIR model having much narrower credible intervals.
The largest discrepancy is that the phenomenological approach model estimates a second peak in mid-January in several regions (most clearly in London, North West, and the East of England).
However, except in London, the credible intervals still overlap.
This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the phenomenological model.
\begin{figure}
    \centering \includegraphics{transmission/compare-regions}
    \caption[Comparing each models estimate of regional incidence.]{%
        Incidence proportion in England and each region estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the phenomenological results.
    }
    \label{transmission:fig:compare-regions}
\end{figure}

@@

The phenomenological method benefits from being able to be implemented in a fast, approximate inference package (R-INLA) that easily scales to many parameters, including random effects.
Therefore, it is far more flexible and can easily include time-varying effects, ethnicity, etc.
The SEIR model is limited by the inference procedure and the requirement to solve an ODE to evaluate the likelihood.
To incorporate a more flexible model, and reduce its bias, a more efficient inference procedure would be required.

Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev,SEIR:fig:prev-appendix}).
This is because the model is not flexible enough to capture the true age distribution of infections.
The overestimation of prevalence in the 12--16 age group and underestimation in the 17--25 age group is reflected when comparing incidence estimates to those from the phenomenological model, which have reduced bias.
\begin{figure}
    \centering \includegraphics{transmission/compare-NE}
    \caption[Comparing each models estimate of North East incidence proportion by age.]{%
        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the phenomenological results.
    }
    \label{transmission:fig:compare-NE}
\end{figure}
